```{r zKnit, echo=F}
zKnit('/key/BES/2013-10-21-mixedDemo/zKnit/2013-10-22', '/key/BES/2013-10-21-mixedDemo/zKnit/2013-10-22')
```

This is a brief introduction to performing longitudinal analyses with R.
The method is described in various sources, perhaps the most approachable is Singer & Willett (2003) from which the first example was derived.

If you haven't installed packages then see Appendix I (below).
To use these examples you will need to install my BES package and several packages from CRAN.
If your `library` statements fail with a "no such package" error then you should try to install the package.

There are two packages in R with which you can perform mixed model analyses,  the `lme` function from the *nlme* package or the `lmer` function with the *lme4* package.
`lme4` is generally preferred because it's presently under active development and tends to be faster.
However, for various pedagogical reasons ([Bates: lmer, p-values and all that](https://stat.ethz.ch/pipermail/r-help/2006-May/094765.html)), `lmer` does not report degrees of freedom or p-values for fixed effects (though as you'll see below there's a way around this).

### Simple example from Singer & Willett

All of the examples in Singer & Willett are coded in Mplus, MLwiN, HLM, SAS, Stata, R, and SPSS at [UCLA](http://www.ats.ucla.edu/stat/r/examples/alda/).

I'll reproduce an example from Chapter 5 (*Treating time more flexibly*) where they examine changes in CES-D by employment status (model C, p 166ff).

They use *nlme*, but I prefer *lme4*.

```{r ALDA1, echo=T}
library(lme4)
unemployment <- read.table("http://www.ats.ucla.edu/stat/r/examples/alda/data/unemployment_pp.txt", header=T, sep=",")

mer = lmer(cesd ~ months * unemp + (months|id), data=unemployment, REML=F)

summary(mer)
```

Part of the convenience of using R is that model coefficients are readily accessible and can be used to generate plots.

Singer & Willett illustrate a brute force method for plotting the results (brute force in the sense that you need to recode the plotting commands if the model changes -- later I'll show a different method that's less sensitive to model alterations).

```{r ALDA1plot, echo=T, h=5, w=5}
months.seq <- seq(0, 14)
fixef.c <- fixef(mer)
unemp.c1 <- fixef.c[[1]] + fixef.c[[2]]*months.seq + fixef.c[[3]] +
            fixef.c[[4]]*months.seq
unemp.c0 <- fixef.c[[1]] + fixef.c[[2]]*months.seq

plot(months.seq, unemp.c1, type="l", ylim=c(5, 20), ylab="CES-D.hat",
     xlab="Months since job loss")
lines(months.seq, unemp.c0, lty=3)
legend(7, 20, c("Unemp = 1", "Unemp = 0"), lty=c(1, 3))
title("Interaction Between Unemp and Time")
```
For simple models we can generally "see" the effects and the directions of association.
For more complex model, particularly with higher order interactions (especially with time), we've found that plotting the "predicted" values is the only way to visualize the effects.

### An example from BLSA

These data are from repeated administrations of the CVLT.
We'll look at changes over time (linear and quadratic) and differences in rates of change between women and men.
BLSA does not code sex explicitly because men have ID<5000.

In this analysis I centered age at 65 and coded explicitly for the quadratic term (unnecessary but easier to comprehend).

```{r CVL1, h=5, w=5, echo=T}
(load(url('http://handls.nih.gov/zBES/2013-10-21-mixedDemo1.rdata')))

cvl$Sex = (cvl$BLSid<5000)
cvl$Sex = factor(cvl$Sex, labels=zQ(Women,Men))
cvl$Age = cvl$Age - 65
cvl$AgeSq = cvl$Age**2
table(cvl$Sex)

library("lme4")

mer = lmer(CVLtca ~ (Age+AgeSq)*Sex + (Age|BLSid), data=cvl, na.action=na.omit)

summary(mer)

b = fixef(mer)

pAge = seq(50, 90)
xAge = pAge - 65
xAgeSq = xAge**2

pW = b[1] + b[2]*xAge + b[3]*xAgeSq
pM = b[1] + b[2]*xAge + b[3]*xAgeSq + b[4] + b[5]*xAge + b[6]*xAgeSq

plot(pAge, pM, type="l", col="blue", ylim=c(25,65), ylab="CVLtca", xlab="Age")
lines(pAge, pW, lty=1, col="red")
legend(75, 65, zQ(Women,Men), lty=c(1,1), col=zQ(red,blue))
```

This method is also "brute force."
The code below yields the same plot using *zMixHat*, a utility function included in the BES package.
I also illustrate a few other commands that make the code a little more accessible and the plot a little prettier.

```{r CVL2, h=5, w=5, echo=T}
library(BES)

pAge = seq(50, 90)
xAge = pAge - 65
xAgeSq = xAge**2

cvlHatCI = zMixHat(cvl, mer, vary="Age=xAge, Sex=zQ(Women,Men)", derivCov="AgeSq=Age**2")
head(cvlHatCI)

par(las=1, lwd=2)
with(cvlHatCI[cvlHatCI$Sex=="Men",],   plot (pAge, hat, type="l", col="blue", ylim=c(25,65), ylab="CVLtca", xlab="Age"))
with(cvlHatCI[cvlHatCI$Sex=="Women",], lines(pAge, hat, lty=1, col="red"))

legend(75, 65, zQ(Women,Men), lty=c(1,1), col=zQ(red,blue))
```
### HANDLS example

Perhaps more relevant but with only one repeated measure we also have longitudinal data in HANDLS.
In this example I examine change over approximately 5 years by poverty status, race, and sex for BVR total errors.
I'll start with a complete model including the (dreaded) 4-way interaction.
I also apply the `cftest` function from the *multcomp* package to generate t-tests and p-values for the fixed effects.

```{r HNDbvr1, echo=T}
library(lme4)
library(multcomp)
library(zStat)
(load(url('http://handls.nih.gov/zBES/2013-10-22-mixedDemo2.rdata')))
describe(BVR)
merBVR1 = lmer(BVRtot ~ PovStat * Race * Sex * Age + (Age|HNDid), data=BVR, na.action=na.omit)
summary(merBVR1)
cftest(merBVR1)
```

Using backward elimination we can remove the 4-way interaction because it was nonsignificant.
Here's a convenient syntactical trick for specifying all of the interactions up to a specific number of terms (note pluses within parentheses and the \^3 indicates the maximum "ways").

```{r HNDbvr2, echo=T, h=7, w=7}
merBVR2 = lmer(BVRtot ~ (PovStat + Race + Sex + Age)^3 + (Age|HNDid), data=BVR, na.action=na.omit)

summary(merBVR2)
cftest(merBVR2)
```

Now this is getting a little more interesting.
It's nearly impossible to get a feel for the direction and magnitude of the effects because the coefficients are not independent (uncorrelated).
Plotting the predicted values by group is an easy way to grasp the overall directions and type of interactions.

```{r HNDbvr3, echo=T, h=7, w=7}
pAge = seq(30, 70)

hatBVR1 = zMixHat(BVR, merBVR2, vary="Age=pAge, PovStat=zQ(abovePovStat,belowPovStat),Race=zQ(White,AfrAm),Sex=zQ(Women,Men)")

par(las=1, lwd=2)

HNDcolors = HNDpltColors()

with(hatBVR1[hatBVR1$Sex=="Women" & hatBVR1$PovStat=='abovePovStat' & hatBVR1$Race=='White',],
	plot (pAge, hat, lty=1, col="red", typ='l', ylim=c(0,14), ylab='BVR', xlab='Age'))
with(hatBVR1[hatBVR1$Sex=="Women" & hatBVR1$PovStat=='abovePovStat' & hatBVR1$Race=='AfrAm',],
	lines(pAge, hat, lty=2, col="red"))
with(hatBVR1[hatBVR1$Sex=="Women" & hatBVR1$PovStat=='belowPovStat' & hatBVR1$Race=='White',],
	lines(pAge, hat, lty=3, col="red"))
with(hatBVR1[hatBVR1$Sex=="Women" & hatBVR1$PovStat=='belowPovStat' & hatBVR1$Race=='AfrAm',],
	lines(pAge, hat, lty=4, col="red"))

with(hatBVR1[hatBVR1$Sex=="Men"   & hatBVR1$PovStat=='abovePovStat' & hatBVR1$Race=='White',],
	lines(pAge, hat, lty=1, col="blue", typ='l', ylim=c(0,14), ylab='BVR', xlab='Age'))
with(hatBVR1[hatBVR1$Sex=="Men"   & hatBVR1$PovStat=='abovePovStat' & hatBVR1$Race=='AfrAm',],
	lines(pAge, hat, lty=2, col="blue"))
with(hatBVR1[hatBVR1$Sex=="Men"   & hatBVR1$PovStat=='belowPovStat' & hatBVR1$Race=='White',],
	lines(pAge, hat, lty=3, col="blue"))
with(hatBVR1[hatBVR1$Sex=="Men"   & hatBVR1$PovStat=='belowPovStat' & hatBVR1$Race=='AfrAm',],
	lines(pAge, hat, lty=4, col="blue"))

legend(30, 14, zQ(abovePovWhite,abovePovAfrAm,belowPovWhite,belowPovAfrAm), lty=1:4, col='black')
text  (30, 10, 'Women in red', adj=c(0,0), col='red')
text  (30, 9.5, 'Men in blue',  adj=c(0,0), col='blue')
```

Although it's a little hard to discern, it appears as though differences associated with poverty status are larger than those associated with race differences.
We can reanalyze and re-plot, or we can recalculate the predicted scores holding constant race (these aren't necessarily equivalent approaches).
Just for illustration, I'll recalculate the predicted scores with `zMixHat` and re-plot.

```{r HNDbvr4, echo=T, h=7, w=7}
par(las=1, lwd=2)

hatBVR2 = zMixHat(BVR, merBVR2, vary="Age=pAge, PovStat=zQ(abovePovStat,belowPovStat),Sex=zQ(Women,Men)", fixedCov='Race')

with(hatBVR2[hatBVR2$Sex=="Women" & hatBVR2$PovStat=='abovePovStat',],
	plot (pAge, hat, lty=1, col="red", typ='l', ylim=c(0,14), ylab='BVR', xlab='Age'))
with(hatBVR2[hatBVR2$Sex=="Women" & hatBVR2$PovStat=='belowPovStat',],
	lines(pAge, hat, lty=2, col="red"))
with(hatBVR2[hatBVR2$Sex=="Men"   & hatBVR2$PovStat=='abovePovStat',],
	lines(pAge, hat, lty=1, col="blue"))
with(hatBVR2[hatBVR2$Sex=="Men"   & hatBVR2$PovStat=='belowPovStat',],
	lines(pAge, hat, lty=2, col="blue"))

legend(30, 14, zQ(abovePovWomen,belowPovWomen,abovePovMen,belowPovMen), lty=c(1,2), col=zQ(red,red,blue,blue))
```

This is a quick plot.
I would probably clean up the legend if I were considering publishing it or using it in a more formal presentation

### Appendix I - Install packages

There is a large and growing larger library of packages in CRAN (http://cran.r-project.org/ -- Comprehensive R Archive Network) that you can install into R (select "packages" under *Software* on the left-hand menu).

To install lme4, issue the command:

`install.packages("lme4")`

To use the package, issue the command:

`library(lme4)`

I also created a small package of useful utility routines called BES.
I'm sharing them with you because I use these functions all the time.
It's easier to share the functions than it is to rewrite all my code.

Install the *BES* package by downloading the file from [BES archive](https://github.com/HANDLS/HANDLS/tree/HNDlocal/R/pkg)
and issuing this *install.package* command:

```
install.packages('BES_0.01.tar.gz', repos=NULL, type='source')
```

### Appendix II - Resources

[Douglas Bates: lme4: Mixed-effects modeling with R](http://lme4.r-forge.r-project.org/lMMwR/lrgprt.pdf)

[John Fox: Linear mixed models](http://cran.r-project.org/doc/contrib/Fox-Companion/appendix-mixed-models.pdf)

[Mixed models FAQ](http://glmm.wikidot.com/faq)

[Venables: Exegeses on Linear Models](http://www.stats.ox.ac.uk/pub/MASS3/Exegeses.pdf)

### References
Singer, J. D., & Willett, J. B. (2003). Applied Longitudinal Data Analysis: Modeling Change and Event Occurrence (1st ed.). Oxford University Press, USA.
