```{r nlz, echo=F}
zKnit('/key/LPS/2013-12-04-R-Basics/test/fig-', '/key/LPS/2013-12-04-R-Basics/test/cache-')
options(digits=10, width=132, xtable.type = 'html')
opts_chunk$set(comment=NA, tidy=F, prompt=T, echo=T)
#
# pandoc -s -S -i -t dzslides --mathjax index.md
# pandoc -s -S -t dzslides --mathjax index.md
# pandoc -s example.md -t latex -o example.pdf
# pandoc -s -S -t dzslides --css test.css --mathjax test.md -o test.htm
# pandoc -s -S -t slidy --mathjax test.md -o test.htm
```

# Some R basics

# General considerations

R takes a programmer's perspective &ndash; makes if difficult for

- all-GUI users
- users unfamiliar with
	- command lines
	- file structures
	- text editors
	- various file formats & translators
	- initialization files (sometimes invisible)
	- operating system locations & conventions

# In this presentation

Focus on

- functional aspect of language
- learn enough to do simple data manipulation & analyses

Some hints

- help is available in R by typing a question mark in front of a function name (e.g., **?summary**)
- typing the name of an object prints the contents of the object
- RStudio may make it easier (particularly as a way to step through a script by executing each line

# Getting data into R

Lots of ways.
We'll mostly use two of them:

- .csv
	- comma-separated values in a text file
	- can create manually or export from a spreadsheet
	- input into R with **read.csv()** function

- .rdata
	- R data file (binary) generally created by R
	- contains all of the variable names, label definitions
	- input into R with **load()** function

# .csv file

Generally looks like

```
column1,column2,column3
1,2,3
2,3,4
3,4,5
```

If these values are in a file called *data/demo1.csv* then you get these data into R by:

```{r}
csvdata = read.csv('data/demo1.csv', stringsAsFactor=FALSE)
csvdata
```

# .rdata file

.rdata files are created by the **save()** function<br>(e.g., **save(csvdata, file='data/demo1.rdata')**)

If we had saved **csvdata** in *data/demo1.rdata* then we can get it into R by:

```{r}
load('data/demo1.rdata')
csvdata
```

# Data in R

Data in R are represented in various ways.

Rather than dwelling on all the forms, for now it is sufficient to think of data in a table of values where the rows are the individuals who were measured and the columns are the names of the measures.

In R this is called a *data frame*.

We can make this concrete with some actual data loaded from the HANDLS website:

```{r}
(load(url('http://handls.nih.gov/zrdata/w00Demographics.rdata')))
```

The function is in parenthesis so that it prints the function's output (the name of the data frame is the output of the **load** function).

# What's in the data frame (head)

**head()** prints the first few lines of data.

```{r}
head(w00Dem)
```
Each row (line) is a HANDLS participant in this example.
The columns are characteristics of the participants (e.g., Age0: initial age; Race, etc).

# What's in the data frame (tail)

**tail()** prints the last few lines of data.

```{r}
tail(w00Dem)
```

Both **head** & **tail** print 5 lines by default.

If want want more or less then include **n=#** where # is the number of lines you want to print (e.g., **tail(w00Dem, n=10)** to print the last 10 lines.

# Descriptive statistics

The **summary()** is used for created a variety of reports.
At its most basic, **summary(dataframe)** provides basic statistics (but not *n*).

```{r}
summary(w00Dem)
```

# Descriptive statistics (continued)

The summary provided by the **summary()** function is not very satisfying.

An alternative is the **describe()** function from the *Hmisc* package (neither installed nor loaded by default).

```{r}
library(Hmisc)
describe(w00Dem[,c('Age0','Sex')])
```

*Hmisc* is an example of a package, a collection of user defined functions available from CRAN, a public archive a R packages.

# Descriptive statistics (continued again)

Another descriptive display is available in the **psych** package, which has a variety of other functions useful for psychometrics
(e.g., component analysis with rotations such as varimax and promax, alpha internal consistency, path diagrams).

```{r}
library(psych)
describe(w00Dem[,3:8])	#remove 1st and 2nd columns (HNDid & DOB) because HNDid isn't meaningful and DOB chokes this function
```

# Merging data frames

- Often data are derived from separate sources and are stored in separate data frames.
- For example, initial demographics are separate from cogntive test scores.
- This prevents duplicating the same information every time we create data.
- Combining data from two separate sources requires a key variable that identifies each line of data
- Generally, we use the HANDLS ID

# Merging (continued)

Merge wave 1 neuropsych data with demographics selecting just age, sex, Benton total errors, & Trails B seconds.

```{r}
(load(url('http://handls.nih.gov/zrdata/w01Neupsy.rdata')))
w01 = merge(w00Dem[,c('HNDid','Age0','Sex')], w01Neupsy[,c('HNDid','BVRtot','TrailsBtestSec')], 'HNDid')
summary(w01)
```

# Selected subsets from a data frame

Separate elements of a data frame are identified as row and column locations.

Rows are identified by

- row numbers
- row name
- logical selectors

Columns are identified by

- column numbers
- column name (variable name)

# Selecting specific rows

Let's say that we want to summarize just the cognitive data for women.

We can do this with the summary command on the merged data frame *w01*.

We find just the women with the logical expression **w01$Sex=='Women'**.

We use this expression in the summary function (or anywhere else) by enclosing it in a **which()** function (the reasons are a bit obscure but are related to what R does when there are missing data).

```{r}
summary(w01[which(w01$Sex=='Women'),])
```

# Correlations

Calculate a correlation matrix.
In the base R package,

```{r}
cor(w01[,c('Age0','BVRtot','TrailsBtestSec')])
```

doesn't seem useful when there are missing data.

# Correlations on missing data (pairwise deletion)

**zCor** from my package *zStat* isn't tripped up by missing data:

```{r}
library(zStat)
zCor(w01[,c('Age0','BVRtot','TrailsBtestSec')])
```

# Descriptive statistics (again)

While we're using *zStat*, you might try:

```{r}
library(zStat)
zDesc(w01)
```

# Descriptive statistics breakdown by group

```{r}
library(doBy)
library(zStat)
summaryBy(BVRtot ~ Sex, w01, FUN=zByFUN)
summaryBy(TrailsBtestSec ~ Sex, w01, FUN=zByFUN)
```

# Linear regression

Although R has routines for t tests (**t.test**), oneway ANOVA (**oneway** in the *lattice* package), and ANOVA (**aov**),
we tend to use general linear model formulations most frequently.

Note the **aov** secretly performs regression and displays the results as a tradional ANOVA table.

A simple multiple regression is done with **lm** &ndash; to regress age, sex, and age × sex on Benton total score:

```{r}
(mod.result1 = lm(BVRtot~Age0*Sex, w01))
```

# Linear regression summary

A more interpretable display is available by applying the summary function to the object holding the linear regression results (**mod.results1**).

```{r}
summary(mod.result1)
```

# Linear regression (just main effects)

For just main effects:

```{r}
mod.result2 = lm(BVRtot~Age0+Sex, w01)
summary(mod.result2)
```

# Comparing models

Although not entirely useful here, we can compare the two linear models to determine whether the interaction term adds useful information.
This command is general in that it will compare models from other types of regression (e.g., nonlinear, mixed-model).

```{r}
anova(mod.result1, mod.result2)
```
Note that in this simple case the probablility for the difference between the two models (0.477) is the same as the probability of t in the regression analysis).

# Categories

Most variables in R are numeric on a continuous scale (e.g., height).

- Generally, this includes counts even though frequencies are integers.
- However, other measures are truly categories (e.g., sex) or categories we create for convenience (e.g., median splits).

In R, these variables are called **factors**.

- Factors are an enormous shorthand.
- R deals with factors appropriately in analyses (e.g., creates suitable dummy coding in regressions).
- But factors are also inconvenient because they do not naturally combine with numeric data (only an example can make this clear).

# Creating factors

In our example data, age is a continous variable with a (more or less) rectangular distribution per the HANDLS sampling design.

```{r hist1,fig.height=5,fig.width=5}
hist(w01$Age0, main='Frequency histogram for baseline age')
```

```{r echo=F}
# edit html -- remove alt & paragraph, add: height='500', width='500'
```

# Factors (continued)

Imagine that we have some reason to believe that there are meaningful differences in Benton errors among three age groups, 30-39, 40-49, & 50-64).

We can create these categories in several ways, perhaps the most elegant of which is

```{r}
w01$age.grp = factor((w01$Age0>=30) + (w01$Age0>=40) + (w01$Age0>=50), levels=1:3, labels=c('30-39', '40-49', '50-64'))
summaryBy(Age0~age.grp, w01, FUN=zByFUN)
```

# Regression models with factors as predictors

Let's see if we were right about age differences in Benton performance.

We might also suspect that age interacts with sex.

---

```{r}
summary(lm(BVRtot ~ age.grp*Sex, w01))
```

Were we right?

# Default contrasts

By default, R uses the first level as the reference (e.g., 40-49 *v.* 30-39, 50-64 *v.* 30-39).

Like most things, there are ways to change this behavior, including the contrast coding (e.g., effect instead of dummy coding).

# What we did and what we didn't

This presentation provides some of the basics but didn't touch on lots of other things

- Logical operators, particularly how to select specific rows and variables
- Presented how to create a factor but not how to convert a factor into a numeric variable
- Manipulating variable names and labels
- Installing packages
- Writing functions

# Assignment symbols & other (quirky) details

- Use "<-" or "="?
- R is case-sensitive
- Though I don't use all of their suggestions, many coders use [Google's R style guide](http://google-styleguide.googlecode.com/svn/trunk/google-r-style.html#assignment)
- R has no line terminator (e.g., ';' is not required at the end of a line)
- Except for a very small number of exceptions, "commands" in R are functions
	- The significance of this is hard to see at first glance
	- The most important aspect of this "feature" is that every function that produces output can also store its output in an object
	- This means that the results of even the most basic functions can be manipulated or output in ways that suit us
	- Functions are used in complicated expressions (e.g., **summary(lm(...)**) in the last slide)

# Some conveniences

- Can define working environment (.Rprofile)
	- load packages (e.g., Hmisc) automatically
	- define options (line width, prompts, etc)
	- works in *R* & *RStudio*
